---
title: "Intro-to-R Workshop Exercise"
author: "ER Deyle"
date: "2023-07-14"
output:
  pdf_document: default
  html_document: default
subtitle: Dealing with Data
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

*Portions of this material were adapted from:*

> François Michonneau, Tracy Teal, Auriel Fournier, Brian Seok, Adam
> Obeng, Aleksandra Natalia Pawlik, ... Ye Li. (2019, July 1).
> datacarpentry/R-ecology-lesson: Data Carpentry: Data Analysis and
> Visualization in R for Ecologists, June 2019 (Version v2019.06.1).
> Zenodo. <http://doi.org/10.5281/zenodo.3264888>

## Purpose

The primary objective of this workshop is to get every BU Marine
Semester student comfortable with using R to do simple tasks that come
up in many courses as you collect, analyze, and visualize data. To this
end, we're going to spend a bit of the time discussing data structures.
R doesn't make this nearly as complicated as some other coding
environments, but ultimately all computational science has to deal with
translating numbers and information we can interpret as scientists into
the sequence of 1's and 0's that computers ultimate must work with. To
get everyone introduced to the basics in R, we're going to go from:

-   a variable (object) with a single value
-   vector objects with multiple values of the same type
-   list objects with multiple values of different types
-   data frame objects that treat multiple lists as columns of a table

While data observations you make (like today the length, width, and
height of gastropod shells) can be stored in the simpler forms, the
functions you may likely use to analyze data (say, perform a linear
regression or analysis of variance) will generally return lists or data
frames.

## Learning Outcomes

By the end of this lesson you should:

-   Know your way around the R-studio "IDE"
-   Know how to use the R Console to do arithmetic
-   How to store values in variables, vectors, lists, and data frames
-   Use these data objects to do basic computations

The first half will involve reading and typing commands into the R
console, as well as doing a few activities. In the second half you will
do a short exercise to put these basic pieces together to do a bit of
marine science and examining the relationship between the length and
weight of gastropod shells that BU Ph.D. student Daniel Wuitchik is
studying for his dissertation.

# Variables and Vectors

## Creating objects in R

```{r, echo = FALSE, purl = TRUE}
### Creating objects in R
```

You can get output from R simply by typing math in the console. In R
studio the `Console` tab should by default be available in the lower
left panel. Type a few simple arithmetic expressions at the `>` and hit
enter/return.

```{r, purl = FALSE}
3 + 5
12 / 7
```

Great! Now you can use R to add decimal numbers instead of googling it.
However, to do useful and interesting things, we need to assign *values*
to *objects*. To create an object, we need to give it a name followed by
the assignment operator `<-`, and the value we want to give it:

```{r, purl = FALSE}
weight_kg <- 55
```

`<-` is the assignment operator. It assigns values on the right to
objects on the left. So, after executing `x <- 3`, the value of `x` is
`3`. For historical reasons, you can also use `=` for assignments, but
not in every context. Because of the
[slight](https://blog.revolutionanalytics.com/2008/12/use-equals-or-arrow-for-assignment.html)
[differences](https://renkun.me/2014/01/28/difference-between-assignment-operators-in-r/)
in syntax, it is good practice to always use `<-` for assignments.

In RStudio, typing <kbd>Alt</kbd> + <kbd>-</kbd> (push <kbd>Alt</kbd> at
the same time as the <kbd>-</kbd> key) will write `<-` in a single
keystroke in a PC, while typing <kbd>Option</kbd> + <kbd>-</kbd> (push
<kbd>Option</kbd> at the same time as the <kbd>-</kbd> key) does the
same in a Mac.

Objects can be given almost any name such as `x`, `current_temperature`,
or `subject_id`. Here are some further guidelines on naming objects:

-   You want your object names to be explicit and not too long.
-   They cannot start with a number (`2x` is not valid, but `x2` is).
-   R is case sensitive, so for example, `weight_kg` is different from
    `Weight_kg`.
-   There are some names that cannot be used because they are the names
    of fundamental functions in R (e.g., `if`, `else`, `for`, see
    [here](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Reserved.html)
    for a complete list). In general, even if it's allowed, it's best to
    not use other function names (e.g., `c`, `T`, `mean`, `data`, `df`,
    `weights`). If in doubt, check the help to see if the name is
    already in use.
-   It's best to avoid dots (`.`) within names. Many function names in R
    itself have them and dots also have a special meaning (methods) in R
    and other programming languages. To avoid confusion, don't include
    dots in names.
-   It is recommended to use nouns for object names and verbs for
    function names.
-   Be consistent in the styling of your code, such as where you put
    spaces, how you name objects, etc. Styles can include "lower_snake",
    "UPPER_SNAKE", "lowerCamelCase", "UpperCamelCase", etc. Using a
    consistent coding style makes your code clearer to read for your
    future self and your collaborators. In R, three popular style guides
    come from [Google](https://google.github.io/styleguide/Rguide.xml),
    [Jean Fan](https://jef.works/R-style-guide/) and the
    [tidyverse](https://style.tidyverse.org/). The tidyverse style is
    very comprehensive and may seem overwhelming at first. You can
    install the [**`lintr`**](https://github.com/jimhester/lintr)
    package to automatically check for issues in the styling of your
    code.

> ### Objects vs. variables
>
> What are known as `objects` in `R` are known as `variables` in many
> other programming languages. Depending on the context, `object` and
> `variable` can have drastically different meanings. However, in this
> lesson, the two words are used synonymously. For more information see:
> <https://cran.r-project.org/doc/manuals/r-release/R-lang.html#Objects>

When assigning a value to an object, R does not print anything. You can
force R to print the value by using parentheses or by typing the object
name:

```{r, purl = FALSE}
weight_kg <- 55    # doesn't print anything
(weight_kg <- 55)  # but putting parenthesis around the call prints the value of `weight_kg`
weight_kg          # and so does typing the name of the object
```

Now that R has `weight_kg` in memory, we can do arithmetic with it. For
instance, we may want to convert this weight into pounds (weight in
pounds is 2.2 times the weight in kg):

```{r, purl = FALSE}
2.2 * weight_kg
```

We can also change an object's value by assigning it a new one:

```{r, purl = FALSE}
weight_kg <- 57.5
2.2 * weight_kg
```

This means that assigning a value to one object does not change the
values of other objects. For example, let's store the animal's weight in
pounds in a new object, `weight_lb`:

```{r, purl = FALSE}
weight_lb <- 2.2 * weight_kg
```

and then change `weight_kg` to 100.

```{r, purl = FALSE}
weight_kg <- 100
```

What do you think is the current content of the object `weight_lb`?
126.5 or 220?

#### ACTIVITY:

> What are the values after each statement in the following?
>
> ```{r, purl = FALSE}
> mass <- 47.5            # mass?
> age  <- 122             # age?
> mass <- mass * 2.0      # mass?
> age  <- age - 20        # age?
> mass_index <- mass/age  # mass_index?
> ```

```{r, echo = FALSE, purl = TRUE}
### Challenge
##
## What are the values after each statement in the following?
##
## mass <- 47.5            # mass?
## age  <- 122             # age?
## mass <- mass * 2.0      # mass?
## age  <- age - 20        # age?
## mass_index <- mass/age  # mass_index?
```

## Functions of variables

Functions are "canned scripts" that automate more complicated sets of
commands including operations, assignments, etc. Many functions are
predefined, or can be made available by importing R *packages* (more on
that later). A function usually takes one or more inputs called
*arguments*. Functions often (but not always) return a *value*. A
typical example would be the function `sqrt()`. The input (the argument)
must be a number, and the return value (in fact, the output) is the
square root of that number. Executing a function ('running it') is
called *calling* the function. An example of a function call is:

```{r, eval = FALSE, purl = FALSE}
weight_kg <- sqrt(10)
```

Here, the value of 10 is given to the `sqrt()` function, the `sqrt()`
function calculates the square root, and returns the value which is then
assigned to the object `weight_kg`. This function takes one argument,
other functions might take several.

The return 'value' of a function need not be numerical (like that of
`sqrt()`), and it also does not need to be a single item: it can be a
set of things, or even a dataset. We'll see that when we read data files
into R.

Arguments can be anything, not only numbers or filenames, but also other
objects. Exactly what each argument means differs per function, and must
be looked up in the documentation (see below). Some functions take
arguments which may either be specified by the user, or, if left out,
take on a *default* value: these are called *options*. Options are
typically used to alter the way the function operates, such as whether
it ignores 'bad values', or what symbol to use in a plot. However, if
you want something specific, you can specify a value of your choice
which will be used instead of the default.

Let's try a function that can take multiple arguments: `round()`.

```{r, results = 'show', purl = FALSE}
round(3.14159)
```

Here, we've called `round()` with just one argument, `3.14159`, and it
has returned the value `3`. That's because the default is to round to
the nearest whole number. If we want more digits we can see how to do
that by getting information about the `round` function. We can use
`args(round)` to find what arguments it takes, or look at the help for
this function using `?round`.

```{r, results = 'show', purl = FALSE}
args(round)
```

```{r, eval = FALSE, purl = FALSE}
?round
```

We see that if we want a different number of digits, we can type
`digits = 2` or however many we want.

```{r, results = 'show', purl = FALSE}
round(3.14159, digits = 2)
```

If you provide the arguments in the exact same order as they are defined
you don't have to name them:

```{r, results = 'show', purl = FALSE}
round(3.14159, 2)
```

And if you do name the arguments, you can switch their order:

```{r, results = 'show', purl = FALSE}
round(digits = 2, x = 3.14159)
```

It's good practice to put the non-optional arguments (like the number
you're rounding) first in your function call, and to then specify the
names of all optional arguments. If you don't, someone reading your code
might have to look up the definition of a function with unfamiliar
arguments to understand what you're doing.

## Vectors and data types

```{r, echo = FALSE, purl = TRUE}
### Vectors and data types
```

A vector is the most common and basic data type in R, and is pretty much
the workhorse of R. A vector is composed by a series of values, which
can be either numbers or characters. We can assign a series of values to
a vector using the `c()` function. For example we can create a vector of
animal weights and assign it to a new object `weight_g`:

```{r, purl = FALSE}
weight_g <- c(50, 60, 65, 82)
weight_g
```

A vector can also contain characters, say if we have categorical
information to store:

```{r, purl = FALSE}
stages <- c("juvenile", "juvenile", "juvenile","adult")
stages
```

The quotes around "juvenile", "adult", etc. are essential here. Without
the quotes R will assume objects have been created called `juvenile` and
`adult`. As these objects don't exist in R's memory, there will be an
error message. [Try it!].

There are many functions that allow you to inspect the content of a
vector. `length()` tells you how many elements are in a particular
vector:

```{r, purl = FALSE}
length(weight_g)
length(stages)
```

An important feature of a vector, is that all of the elements are the
same type of data. The function `class()` indicates what kind of object
you are working with:

```{r, purl = FALSE}
class(weight_g)
class(stages)
```

The function `str()` provides an overview of the structure of an object
and its elements. It is a useful function when working with large and
complex objects:

```{r, purl = FALSE}
str(weight_g)
str(stages)
```

You can use the `c()` function to add other elements to your vector:

```{r, purl = FALSE}
weight_g <- c(weight_g, 90) # add to the end of the vector
weight_g <- c(30, weight_g) # add to the beginning of the vector
weight_g
```

In the first line, we take the original vector `weight_g`, add the value
`90` to the end of it, and save the result back into `weight_g`. Then we
add the value `30` to the beginning, again saving the result back into
`weight_g`.

We can do this over and over again to grow a vector, or assemble a
dataset. As we program, this may be useful to add results that we are
collecting or calculating.

An **atomic vector** is the simplest R **data type** and is a linear
vector of a single type. Above, we saw 2 of the 6 main **atomic vector**
types that R uses: `"character"` and `"numeric"` (or `"double"`). These
are the basic building blocks that all R objects are built from. The
other 4 **atomic vector** types are:

-   `"logical"` for `TRUE` and `FALSE` (the boolean data type)
-   `"integer"` for integer numbers (e.g., `2L`, the `L` indicates to R
    that it's an integer)
-   `"complex"` to represent complex numbers with real and imaginary
    parts (e.g., `1 + 4i`) and that's all we're going to say about them
-   `"raw"` for bitstreams that we won't discuss further

You can check the type of your vector using the `typeof()` function and
inputting your vector as the argument.

Vectors are one of the many **data structures** that R uses. Other
important ones are lists (`list`), matrices (`matrix`), data frames
(`data.frame`), factors (`factor`) and arrays (`array`).

#### ACTIVITY:

> ### Challenge
>
> -   We've seen that atomic vectors can be of type character, numeric
>     (or double), integer, and logical. But what happens if we try to
>     mix these types in a single vector?
>
> ```{text_answer, echo=FALSE, purl=FALSE}
> R implicitly converts them to all be the same type
> ```
>
> -   What will happen in each of these examples? (hint: use `class()`
>     to check the data type of your objects):
>
>     ``` r
>     num_char <- c(1, 2, 3, "a")
>     num_logical <- c(1, 2, 3, TRUE)
>     char_logical <- c("a", "b", "c", TRUE)
>     tricky <- c(1, 2, 3, "4")
>     ```
>
> -   Why do you think it happens?
>
> ```{text_answer, echo=FALSE, purl=FALSE}
> Vectors can be of only one data type. R tries to
> convert (coerce) the content of this vector to find a "common
> denominator" that doesn't lose any information.
> ```
>
> -   How many values in `combined_logical` are `"TRUE"` (as a
>     character) in the following example (reusing the 2 `..._logical`s
>     from above):
>
>     ``` r
>     combined_logical <- c(num_logical, char_logical)
>     ```
>
> ```{text_answer, echo=FALSE, purl=FALSE}
> Only one. There is no memory of past data types, and the coercion happens the
> first time the vector is evaluated. Therefore, the `TRUE` in `num_logical`
> gets converted into a `1` before it gets converted into `"1"` in
> `combined_logical`.
> ```
>
> -   You've probably noticed that objects of different types get
>     converted into a single, shared type within a vector. In R, we
>     call converting objects from one class into another class
>     *coercion*. These conversions happen according to a hierarchy,
>     whereby some types get preferentially coerced into other types.
>     Can you draw a diagram that represents the hierarchy of how these
>     data types are coerced?
>
> ```{text_answer, echo=FALSE, purl=FALSE}
> logical &#8594; numeric &#8594; character &#8592; logical
> ```

```{r, echo = FALSE, eval = FALSE, purl = TRUE}
## We’ve seen that atomic vectors can be of type character, numeric, integer, and
## logical. But what happens if we try to mix these types in a single
## vector?

## What will happen in each of these examples? (hint: use `class()` to
## check the data type of your object)
num_char <- c(1, 2, 3, "a")

num_logical <- c(1, 2, 3, TRUE)

char_logical <- c("a", "b", "c", TRUE)

tricky <- c(1, 2, 3, "4")

## Why do you think it happens?


## How many values in `combined_logical` are `"TRUE"` (as a character) in the
## following example:
combined_logical <- c(num_logical, char_logical)

## You've probably noticed that objects of different types get
## converted into a single, shared type within a vector. In R, we call
## converting objects from one class into another class
## _coercion_. These conversions happen according to a hierarchy,
## whereby some types get preferentially coerced into other types. Can
## you draw a diagram that represents the hierarchy of how these data
## types are coerced?
```

## Subsetting vectors

If we want to extract one or several values from a vector, we must
provide one or several indices in square brackets. For instance:

```{r, results = 'show', purl = FALSE}
fish <- c("hake", "herring", "sprat", "cod")
fish[2]
fish[c(3, 2)]
```

We can also repeat the indices to create an object with more elements
than the original one:

```{r, results = 'show', purl = FALSE}
more_fish <- fish[c(1, 2, 3, 2, 1, 4)]
more_fish
```

R indices start at 1. Programming languages like Fortran, MATLAB, Julia,
and R start counting at 1, because that's what human beings typically
do. Languages in the C family (including C++, Java, Perl, and Python)
count from 0 because that's simpler for computers to do.

## List objects

Lists are a more general and flexible data object that are otherwise
very similar to vectors. Why are they more flexible? Well, above we
showed how a vector couldn't contain values of different classes. In a
list that restriction is gone. Let's have a look.

First, you can create a list object much the same way as a vector, but
instead of wrapping the list of values within `c()` instead you use
`list()`.

```{r}
num_logical <- list(1, 2, 3, TRUE)
```

This new list has the same length it would as a vector.

```{r}
length(num_logical)
```

But now when we ask about the class, what do we get?

```{r}
class(num_logical)
```

Interesting.

`[ ]` selects a list of the elements while `[[ ]]` selects the single
element of the list.

Finally, interacting with list data can be especially easy when you give
names.

```{r}
specimen_1 <- list(common_name="bay scallop",width_in=4.25,height_in=0.5,length_in=2.625,sex="female")
```

We can refer to list element by name using the `$` between the list
object name and the name of the element we want inside.

```{r}
specimen_1[[3]]
specimen_1$height_in
```

Have we done anything yet that we couldn't do with vectors? Not really.
Here's the kicker, though, not only can the list elements be a mix of
numbers, logicals, and characters, but the list elements can also be
vectors... or even more lists!

```{r}
specimen_data <- list(weight_g = c(50, 60, 65, 82),
                      stages=c("juvenile","juvenile","juvenile","adult"))
```

Why do you need to know how to use lists? A primary concern for this
workshop is that if you run most statistical tests in R, be it an ANOVA
or a linear regression, those functions tend to return list objects
since the analysis results contain multiple kinds of information. For
example, if we just grab one of the included R datasets, `trees` (type
`?trees` in console if you want a bit of context), we can run a simple
linear regression with `lm` (type `?lm` while you're at it!).

```{r}
lm_out <- lm(Volume ~ Girth,data=trees)
```

We save the output of the `lm` function call. If you look in the
**Environment** tab in the upper right of R-studio you'll see
`List of 12`. We can check for ourselves:

```{r}
length(lm_out)
names(lm_out)
```

Lots of different information here, including the linear model
coefficients, significance levels, and residuals. The `coefficients`
element of the list happens to be a numeric named vector:

```{r}
str(lm_out$coefficients)
```

While the `model` element is something called a "data.frame" (more on
that in just a minute).

```{r}
str(lm_out$model)
```

## Data frame objects

Now we come to data frames or `data.frames`. Data frames are really just
special kinds of lists that make dealing with multivariate data much
easier and breazier by assuming some constraints. Basically, a data
frame is a list of variables vectors that all have the same length.
There's lots of tools you can pick up along the way to work with
data.frame objects, but doing the basic interfacing as if it were a list
can get you a long way.

```{r}
str(trees)
```

If you want to extract the girth of the 5th tree for example:

```{r}
trees$Girth[5]
```

The data frame can also act like an array, which we've skipped over a
bit, but is the generalization of a vector that has rows and columns of
data. Since Girth is the first variable, we can pull the same value by
requesting the 5th row and the 1st column:

```{r}
trees[5,1]
```

Finally, it can be useful to add on new columns of data as you do
calculations. For example, if you read the help entry on the `trees`
data set, you might have seen the note that the variable `Girth` is an
erroneous labeling of diameter data. Girth really is more suggestive of
circumference than diameter. But we could do that conversion! Assuming
the trunk has a roughly circular cross-section, then
`circumference = pi * diameter`. And guess what, R knows the value of
pi! (Well a lot of the digits at least).

First, let us rename `Girth`. We can do this via the `names()` function.

```{r}
names(trees)
names(trees)[1] <- "Diameter"
names(trees)
```

Now what happens if we ask for the "Girth"?

```{r}
trees$Girth
```

It's empty! But, R is more than happy to let us assign something there.

```{r}
trees$Girth <- pi * trees$Diameter
trees$Girth
```

Great! So now you have an idea how to interact with data.frames,
including renaming variables and construct additional columns. Why do
you need to know how to use data.frames? As we've already seen, they are
a way to supply data to functions that perform statistical tests. They
are also the way data will often be loaded into R, e.g. if you are
reading in table from an excel or google spreadsheet. And... that is how
we will proceed today!

------------------------------------------------------------------------

------------------------------------------------------------------------

## Putting it all together

At this point you should have the basic idea for how to create variables
in R and perform basic arithmetic. Now let's put it all together to do
some marine science in the form of a paired-down lab exercise. You will
collect data on the length, width, and height of slipper shells, use a few
mathematical formula (dare we call it a model!) to make predictions about
the weight of those shells, then weigh the shells to validate the model
predictions.

### Open a new R script

Just like in other labs, it's good to do your work in some sort of
notebook. Here, we will keep it simple and make a blank R-script in
Rstudio. While we can do all the calculations entering line-by-line into
the console, writing the whole sequence of steps down in order minimizes
error and maximizes the ease of reproducing. In some Marine Semester
classes, you may be introduced to a more elegant notebook format with
Rmarkdown.

There are a few ways to get a new script open in RStudio. There's a
small green "+" sign button in the upper left you can use or you can
navigate to the file menu and select `New File -> R Script`. Your new,
blank file should appear in the upper left pane of RStudio. Go ahead and
save it in your documents folder (or where-ever you prefer) and use a
name like "R-intro-workshop_exercise.R".

### Collect and record data

Courtesy of Dan Wuitchik, there are several sets of slipper shells available
for you to measure. The samples are organized by collection site (Cape May, NJ to
the tip of Maine) and need to be kept that way. Work with other students to measure
the height, width, and length in cm

As you measure, you need to record the measurements. As always will be
the case, there's more than one option. Since everyone should be able to
quickly access Google Sheets, let's do it that way. Many of you will
have collected measurements in a lab into a data sheet; ask for help if
this part isn't obvious. Set the document up like so:

| Length | Width | Height |
|--------|-------|--------|
| \#     | \#    | \#     |
| \#     | \#    | \#     |

#### TASK:

> Record physical measurements of slipper shells in a Google Sheet.

When you are done, we need to export the data so that it's easy to turn
around and load into R.


#### TASK:

> Download the data in your Google Sheet as a "csv" file.


### Import the data into R

There are many functions to help you read data
saved in a file into an object in R from various formats (including .xls
and .xlsx). We're going with the classic, `read.table()`. Have at least
a quick look at the help function by typing:

```{r}
?read.table
```

This is a very flexible function, so it may be overwhelming. It has 25
formal arguments! Luckily, there are sensible default values available
for all but one, `file`. So you'll have to give it the file name. The
other important arguments here are `header` and `sep`. The first of
those specifies if the data coming in has headers (it should) and what
character is uses to separate the values in a row. We've made a "csv",
so the separator is `","`.

```{r,eval=FALSE}
data_shell_sizes <- read.table(file="myfile.csv",header=TRUE,sep=",")
```

#### TASK:

> Use `read.table()` to import your recorded data.

Alas! We have measured in millimeters but I'm going to ask you next to work
in centimeters. The unit conversion for mm to cm is 10 mm = 1 cm.

#### TASK:

> Convert the measurements from inches into centimeters and store in a
> second data.frame.

Now that we have centimeters, we can calculate volume in cubic
centimeters. To do this, you'll need to make some approximations about
the shape of the "specimens". Here are a few formula and pick the one(s)
you think are most appropriate. Real shells are of course neither of
these idealized shapes, but let's see how close we can get. After all,
*all models are wrong, some are useful.*

| Shape          | Formula                     |
|----------------|-----------------------------|
| Disc           | pi \* r\^2 \* h             |
| Half-Ellipsoid | 2/3 pi \* l/2 \* w/2 \* h/2 |

Recall: R knows the approximate value of pi:

```{r}
pi
```

While you do it, go ahead and make use of R data structures and descriptive variable
names! If you come back a week later and just see

```{r}
x1 <- 4.3
x2 <- 1.2
x3 <- 2.9
y <- x1*x2*x3
```

You stand a good chance at not actually remembering what that code was even about.



#### TASK:

> Use arithmetic to estimate the volume of each specimen you measured in
> cubic centimeters. Store the estimated volume as a new column in the data frame
> with your imported observations of length, width, and height.


Aragonite has a density of 2.93 g/cc. However, mollusk shells aren't composed of pure aragonite 
and are also porous, so they are less dense. Let's assume the shells have a density of 2.7 g/cc. 
Use this to estimate the mass of each specimen.

#### TASK:

> Use this approximate density and the estimated volume of each specimen
> to estimate their mass. Again, store this estimated mass as a new column
> in the same data frame.



#### TASK:

> Compare your results to  direct measurements of mass for each
specimen. Collect this new data in the same spreadsheet, adding another column,
then display the data in a graph (see below). How close are the
estimates? Does there appear to be a systematic bias?

HINT: You can create a basic x-y scatter plot in R using the command
`plot(x,y)` where `x` and `y` are vectors of the same length. If you
want to dress it up a bit you can add a "1-1" line afterwards using
`abline(a=0,b=1)`. Type `?abline` to see the syntax. You can pass
additional arguments to fancy it up, like a dotted line type with
`lty=2` and a red color with `color='red'`.


## If you have time

In fisheries, it is often much easier to measure length of individuals than many other related variables of interest like age or weight. Allometric scaling relationships are a way to parameterize relationships between variables to estimate unmeasured variables from length (or in other cases, mass). Let's look at the relationship between length and mass for the shells.

#### TASK:

> Create graph that shows observed mass as a function of length for the specimens you measured, possibly including
> measurement data from other students if you like.
> Does mass look to be predictable from just length? Do you expect this relationship to be linear? Does it appear linear?

## Reflection Questions

-   Was there a systematic bias to your estimates of mass? List the
    different assumptions made along the way to get that estimate. Which
    of these might have been sources of bias? How might you change these
    assumptions to get better estimates? [There is possibly a lesson
    here that correctly written code doesn't necessarily produce correct
    answers!].
-   What questions do you still have about the basics of variables and
    vectors in R?

## Appendix: Resources

More on Crepidula fornicata:

Henry, J. Q., & Lyons, D. C. (2016). Molluscan models: Crepidula fornicata. Current Opinion in Genetics & Development, 39, 138-148. https://doi.org/10.1016/j.gde.2016.05.021

Global change dimensions of mollusk shell properties:

Gizzi, F., Caccia, M., Simoncini, G. et al. Shell properties of commercial clam Chamelea gallina are influenced by temperature and solar radiation along a wide latitudinal gradient. Sci Rep 6, 36420 (2016). https://doi.org/10.1038/srep36420

